---
title: "Kriging in Sports"
subtitle: "CMDA-4654 Project 2"
author: "Group 1 | Baker Dean, Toshaani Dewan?, Nick Fitti, Nico Ottobre, Margaret Smith"
date: "April 2026"
fontsize: 9pt
classoption: "aspectratio=169"
output: 
  beamer_presentation: 
    fonttheme: professionalfonts
    highlight: kate
    theme: Rochester
urlcolor: blue
header-includes:
- \definecolor{VTmaroon}{HTML}{861F41}
- \usecolortheme[named=VTmaroon]{structure}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{bbm}
- \usepackage{cancel}
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{bm}
- \usepackage{enumitem}
- \usepackage{underscore}
---
```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(hoopR)
library(progressr)
library(gstat)
library(sp)
library(automap)
library(patchwork)
library(scales)

knitr::opts_chunk$set(echo = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")

def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  paste0("\n \\", "footnotesize","\n\n", x, "\n\n \\normalsize")
})
```

```{r load-data, cache=TRUE}
# All data loading and function definitions here
# cache=TRUE means this only runs once
# Read in data using hoopR
pbp <- load_nba_pbp(seasons = 2026)
player_box <- load_nba_player_box(seasons = 2026)

player_names <- player_box |>
  distinct(athlete_id, athlete_display_name)
```

```{r}
# Unprinted Code for NBA Example

# Libraries
library(dplyr)
library(hoopR)
library(progressr)
library(ggplot2)
library(sportyR)
library(gstat)
library(sp)
library(automap)

# Function to draft the lines of a basketball halfcourt
draw_court <- function() {
  rim_x <- 5.25
  three_pt_r <- 23.75
  corner_x <- rim_x + sqrt(three_pt_r^2 - 22^2)
  angle_top <- atan2(22, corner_x - rim_x)
  angle_bot <- atan2(-22, corner_x - rim_x)
  
  arc_points <- data.frame(
    x = rim_x + three_pt_r * cos(seq(angle_top, angle_bot, length.out = 1000)),
    y = three_pt_r * sin(seq(angle_top, angle_bot, length.out = 1000))
  )
  
  list(
    # Three point arc
    geom_path(data = arc_points,
              aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # Corner three lines
    geom_path(data = data.frame(x = c(0, corner_x), y = c(22, 22)),
              aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    geom_path(data = data.frame(x = c(0, corner_x), y = c(-22, -22)),
              aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # Paint
    geom_path(data = data.frame(x = c(0, 19, 19, 0, 0), y = c(8, 8, -8, -8, 8)),
              aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # FT circle solid half
    geom_path(data = data.frame(
      x = 19 + 6 * cos(seq(-pi/2, pi/2, length.out = 500)),
      y = 6 * sin(seq(-pi/2, pi/2, length.out = 500))),
      aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # FT circle dashed half
    geom_path(data = data.frame(
      x = 19 + 6 * cos(seq(pi/2, 3*pi/2, length.out = 500)),
      y = 6 * sin(seq(pi/2, 3*pi/2, length.out = 500))),
      aes(x = x, y = y), inherit.aes = FALSE, color = "black",
      linewidth = 0.6, linetype = "dashed"),
    # Restricted arc
    geom_path(data = data.frame(
      x = rim_x + 4 * cos(seq(-pi/2, pi/2, length.out = 500)),
      y = 4 * sin(seq(-pi/2, pi/2, length.out = 500))),
      aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # Rim
    geom_path(data = data.frame(
      x = rim_x + 0.75 * cos(seq(0, 2*pi, length.out = 500)),
      y = 0.75 * sin(seq(0, 2*pi, length.out = 500))),
      aes(x = x, y = y), inherit.aes = FALSE, color = "darkorange2", linewidth = 0.8))
}

# Data Cleaning
shots <- pbp |>
  filter(shooting_play == TRUE,
         !is.na(coordinate_x),
         !is.na(coordinate_y),
         !grepl("Free Throw", type_text),
         season_type == 2) |>
  select(athlete_id_1, coordinate_x, coordinate_y, score_value) |>
  left_join(player_names, by = c("athlete_id_1" = "athlete_id")) |>
  mutate(
    x = ifelse(coordinate_x > 0,
               -(coordinate_x - 46.75),
               coordinate_x + 46.75),
    y = ifelse(coordinate_x > 0, -coordinate_y, coordinate_y)) |>
  filter(x <= 45) |>
  select(athlete_display_name, score_value, x, y)
```

## Data Introduction

Points per shot is one of the biggest current topics of discussion in basketball analytics. It is calculated using two factors: a team or players chance of making a shot from a position on the floor and the point of value of a make from that position. Performing analysis on this topic requires gathering both of those values which can be done using the `hoopr` package.

```{r, echo=TRUE, eval=FALSE}
library(hoopR)

pbp <- load_nba_pbp(seasons = 2026)
player_box <- load_nba_player_box(seasons = 2026)

player_names <- player_box |>
  distinct(athlete_id, athlete_display_name)

shots <- pbp |>
  filter(shooting_play == TRUE,
         !is.na(coordinate_x), !is.na(coordinate_y), !grepl("Free Throw", type_text),
         season_type == 2) |>
  select(athlete_id_1, coordinate_x, coordinate_y, score_value) |>
  left_join(player_names, by = c("athlete_id_1" = "athlete_id")) |>
  mutate(
    x = ifelse(coordinate_x > 0,
               -(coordinate_x - 46.75),
               coordinate_x + 46.75),
    y = ifelse(coordinate_x > 0, -coordinate_y, coordinate_y)) |>
  filter(x <= 45) |>
  select(athlete_display_name, score_value, x, y)
```

## Data Example

The data has been loaded and saved as csv for ease of access: \href{https://docs.google.com/spreadsheets/d/1uBDuThoKEiOK03TE7cipW_oFBo6iOxBon5RDeMG9lUI/edit?usp=sharing}{Click here} to access shots.csv. 
\vspace{0.15cm}
\begin{columns}
\begin{column}{0.45\textwidth}

Preview of the data
\vspace{0.1cm}
```{r}
shots |> 
  head(20) |>
  knitr::kable(format = "latex", booktabs = TRUE)
```

\end{column}
\begin{column}{0.55\textwidth}

NBA Regular Season Shot Locations (2025-26):
\vspace{0.1cm}
```{r, fig.width=4, fig.height=2.33}
ggplot(shots, aes(x = x, y = y)) +
  stat_density_2d(
    aes(fill = after_stat(density)),
    geom = "raster",
    contour = FALSE,
    alpha = 0.9
  ) +
  draw_court() +
  scale_fill_viridis_c() +
  coord_fixed(
    xlim = c(0, 45),
    ylim = c(-25, 25),
    expand = FALSE,
    clip = "off"
  ) +
  labs(
    title = "NBA Shot Locations",
    x = "Distance from Basket (ft)",
    y = "Court Width (ft)",
    fill = "Density"
  ) +
  theme_minimal(base_size = 7) + 
  theme(panel.grid = element_blank(),
      plot.margin = margin(t = 2, r = 0, b = 0, l = -35))
```
\end{column}
\end{columns}

## Kriging Motivation

It is clear from the shot density layout on the previous slide that there are two main areas that teams take shots from:

* the three point line
* the rim

This makes sense given that shots at the rim are likely to be high percentage and shots from three point range are worth more points. The middle area between these two hot zones is fairly empty, and the league has been trending to make midrange jumpshots even less a part of offenses in the Modern Era. All of this evolution is in the pursuit of offensive effiency which is where Kriging comes in.

Since the shot map is made up of x and y data, it can be treated similarly to a geostatistics field from more typical Kriging -  it has high density areas and low gaps with a large number of very individualized spots. Kriging can be used to smooth over these less commonly attempted areas and give a picture of not only where teams shoot but how well they perform from these areas.

Combining those two areas will mean that Kriging can make a points per shot estimation from every location on the floor which can then be used to check if teams are running their offenses to peak effiency.

## Attempt Frequency Kriging

A typical picture of attempt frequency for a basketball team would consist of colored boxes or hexes that represent volume from each area of the court, but a custom Kriging can smooth this out into a continuous map using two foot by two foot bins.

```{r plot-attempts}
team_shots <- pbp |>
  filter(shooting_play == TRUE,
         !is.na(coordinate_x),
         !is.na(coordinate_y),
         !grepl("Free Throw", type_text),
         season_type == 2) |>
  mutate(
    team_name = ifelse(team_id == home_team_id,
                       paste(home_team_name, home_team_mascot),
                       paste(away_team_name, away_team_mascot)),
    x = ifelse(coordinate_x > 0,
               -(coordinate_x - 46.75),
               coordinate_x + 46.75),
    y = ifelse(coordinate_x > 0, -coordinate_y, coordinate_y)
  ) |>
  filter(x <= 45) |>
  select(team_name, score_value, x, y)

plot_attempts <- function(team) {
  
  rim_x <- 5.25
  three_pt_r <- 23.75
  bin_size <- 2
  min_attempts <- 10
  
  # Prediction grid
  pred_grid <- expand.grid(
    x = seq(0.5, 45.5, by = 1),
    y = seq(-24.5, 25.5, by = 1)
  ) |>
    as.data.frame() |>
    mutate(dist_from_rim = sqrt((x - rim_x)^2 + y^2)) |>
    filter(dist_from_rim <= three_pt_r + 5) |>
    select(x, y) |>
    as.data.frame()
  
  coordinates(pred_grid) <- ~x+y
  gridded(pred_grid) <- TRUE
  
  # Bin shots
  attempt_bins <- team_shots |>
    filter(team_name == team) |>
    mutate(
      x_bin = floor(x / bin_size) * bin_size + bin_size / 2,
      y_bin = floor(y / bin_size) * bin_size + bin_size / 2
    ) |>
    group_by(x_bin, y_bin) |>
    summarise(attempts = n(), .groups = "drop") |>
    filter(attempts >= min_attempts) |>
    as.data.frame()
  
  coordinates(attempt_bins) <- ~x_bin+y_bin
  
  # Krige
  krig_attempts_auto <- suppressWarnings(
    autoKrige(attempts ~ 1,
              input_data = attempt_bins,
              new_data = pred_grid)
  )
  
  krig_attempts_df <- as.data.frame(krig_attempts_auto$krige_output)
  names(krig_attempts_df) <- c("x", "y", "attempts_pred", 
                                "attempts_var", "attempts_stdev")
  krig_attempts_df <- krig_attempts_df |>
    mutate(attempts_pred = pmax(0, attempts_pred))
  
  # Plot
  ggplot(krig_attempts_df, aes(x = x, y = y, fill = attempts_pred)) +
    geom_raster(interpolate = TRUE) +
    draw_court() +
    scale_fill_viridis_c() +
    coord_fixed(xlim = c(0, 35), ylim = c(-25, 25)) +
    labs(title = paste(team, "- Shot Attempts via Kriging"),
         x = "Distance from Basket (ft)",
         y = "Court Width (ft)",
         fill = "Attempts") +
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(),
      plot.margin = margin(t = 2, r = 0, b = 0, l = -3))
}
```
\begin{columns}
\begin{column}{0.5\textwidth}

```{r bos-attempts-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_attempts("Boston Celtics")
```

\end{column}
\begin{column}{0.5\textwidth}

```{r det-attempts-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_attempts("Detroit Pistons")
```

\end{column}
\end{columns}

## FG\% Kriging

While the Kriging for shot attempts is somewhat trivial, it is extremely useful for field goal percentage. This statistic leads to a map that has continuous values across all the bins and the maps will differ more greatly team-to-team.

```{r plot-percentage}
plot_fg_pct <- function(team) {
  
  rim_x <- 5.25
  three_pt_r <- 23.75
  bin_size <- 2
  min_attempts <- 10
  
  # Prediction grid
  pred_grid <- expand.grid(
    x = seq(0.5, 45.5, by = 1),
    y = seq(-24.5, 25.5, by = 1)
  ) |>
    as.data.frame() |>
    mutate(dist_from_rim = sqrt((x - rim_x)^2 + y^2)) |>
    filter(dist_from_rim <= three_pt_r + 5) |>
    select(x, y) |>
    as.data.frame()
  
  coordinates(pred_grid) <- ~x+y
  gridded(pred_grid) <- TRUE
  
  # Bin shots
  fg_bins <- team_shots |>
    filter(team_name == team) |>
    mutate(
      x_bin = floor(x / bin_size) * bin_size + bin_size / 2,
      y_bin = floor(y / bin_size) * bin_size + bin_size / 2
    ) |>
    group_by(x_bin, y_bin) |>
    summarise(
      attempts = n(),
      makes = sum(score_value > 0),
      fg_pct = makes / attempts,
      .groups = "drop"
    ) |>
    filter(attempts >= min_attempts) |>
    as.data.frame()
  
  coordinates(fg_bins) <- ~x_bin+y_bin
  
  # Krige
  krig_fg_auto <- suppressMessages(suppressWarnings(
    autoKrige(fg_pct ~ 1,
              input_data = fg_bins,
              new_data = pred_grid)
  ))
  
  krig_fg_df <- as.data.frame(krig_fg_auto$krige_output)
  names(krig_fg_df) <- c("x", "y", "fg_pred", "fg_var", "fg_stdev")
  
  # Return
  return(list(plot = ggplot(krig_fg_df, aes(x = x, y = y, fill = fg_pred)) +
    geom_raster(interpolate = TRUE) +
    draw_court() +
    scale_fill_viridis_c(labels = scales::percent,
                         limits = c(0.20, 0.60),
                         oob = scales::squish) +
    coord_fixed(xlim = c(0, 35), ylim = c(-25, 25)) +
    labs(title = paste(team, "- FG% via Kriging"),
         x = "Distance from Basket (ft)",
         y = "Court Width (ft)",
         fill = "FG%") +
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(),
      plot.margin = margin(t = 2, r = 0, b = 0, l = -3)), data = krig_fg_df))
}
```
\begin{columns}
\begin{column}{0.5\textwidth}

```{r bos-percentage-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
bos_fg <- plot_fg_pct("Boston Celtics")
bos_fg$plot
```

\end{column}
\begin{column}{0.5\textwidth}

```{r det-percentage-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
det_fg <- plot_fg_pct("Detroit Pistons")
det_fg$plot
```

\end{column}
\end{columns}

## Points Per Shot

As mentioned previously, points per shot is a metric that is used to determine the quality of a type of shot attempt. Now that the field goal percentage Kriging has supplied a likelihood of making a shot at any location on the floor, it is easy to take it a step further into point values of the shots.

```{r}
plot_pps <- function(fg_data) {
  
  rim_x <- 5.25
  three_pt_r <- 23.75
  corner_x <- rim_x + sqrt(three_pt_r^2 - 22^2)
  
  pps_df <- fg_data |>
    mutate(
      dist_from_rim = sqrt((x - rim_x)^2 + y^2),
      is_three = dist_from_rim >= three_pt_r | (abs(y) >= 22 & x <= corner_x),
      point_value = ifelse(is_three, 3, 2),
      pps = fg_pred * point_value
    )
  
  ggplot(pps_df, aes(x = x, y = y, fill = pps)) +
    geom_raster(interpolate = TRUE) +
    draw_court() +
    scale_fill_viridis_c(limits = c(0.5, 2.0),
                         oob = scales::squish) +
    coord_fixed(xlim = c(0, 35), ylim = c(-25, 25)) +
    labs(title = "Points Per Shot via Kriging",
         x = "Distance from Basket (ft)",
         y = "Court Width (ft)",
         fill = "PPS") +
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(),
      plot.margin = margin(t = 2, r = 0, b = 0, l = -3))
}
```
\begin{columns}
\begin{column}{0.5\textwidth}

```{r bos-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(bos_fg$data)
```

\end{column}
\begin{column}{0.5\textwidth}

```{r det-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(det_fg$data)
```

\end{column}
\end{columns}

## Non-Kriging Approach

So the question becomes "How do these smooth maps help NBA teams improve their offenses?"

One of the main advantages that these plots provide is the way they make it much easier to full picture of slots across the floor. As a comparison, here would be a more standard look at a teams percentages and shot volumes from Synergy Sports. 

\begin{columns}
\begin{column}{0.5\textwidth}
The plot makes it hard tell whether there's smaller deviations within the large groups, doesn't calculate the point values of the shots (i.e. shooting 50\% from 2 is the same as 33\% from 3).
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=\linewidth]{"C:/Users/bwdea/OneDrive/Pictures/Screenshots/Screenshot 2026-04-21 173823.png"}
\end{column}
\end{columns}

## Kriging Comparison

## Accuracy of Predictions

## Conclusions \& Kriging Limitations

## References