---
title: "Kriging in Sports"
subtitle: "CMDA 4654 Project 2 - Group 1"
author: "Baker Dean, Toshaani Dewan, Nick Fitti, Nico Ottobre, Margaret Smith"
date: "April 2026"
fontsize: 9pt
classoption: "aspectratio=169"
output: 
  beamer_presentation: 
    fonttheme: professionalfonts
    highlight: kate
    theme: Rochester
    citation_package: biblatex
urlcolor: blue
bibliography: reference.bib
biblatexoptions: [backend=biber]
header-includes:
- \definecolor{VTmaroon}{HTML}{861F41}
- \usecolortheme[named=VTmaroon]{structure}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{bbm}
- \usepackage{cancel}
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{bm}
- \usepackage{enumitem}
- \usepackage{underscore}
- \usepackage{biblatex}
- \setlist[itemize]{label=\textbullet}
---
```{r setup, include=FALSE}
#Libraries
library(knitr)
library(dplyr)
library(ggplot2)
library(hoopR)
library(progressr)
library(gstat)
library(sp)
library(automap)
library(patchwork)
library(scales)
library(sportyR)

# Setup
knitr::opts_chunk$set(echo = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")

def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  paste0("\n \\", "footnotesize","\n\n", x, "\n\n \\normalsize")
})
```
## Spatial / Geostatistical Data

\vspace{0.35cm}

\textbf{What is spatial data?}

\begin{itemize}
\item Observations collected at specific locations
\item Each point has coordinates $(x,y)$ and a measured value $Z(x)$
\item Location matters because nearby points often behave similarly
\item Values may change across space rather than randomly
\end{itemize}


\textbf{Key Idea: Spatial Dependence}

\begin{itemize}
\item Points close together tend to have more similar values
\item Points farther apart tend to be less related
\item This relationship can be modeled statistically
\end{itemize}

\vspace{0.3cm}

\centering
\Large
Nearby $\Rightarrow$ Similar

\vspace{0.25cm}

Farther Away $\Rightarrow$ Less Similar

\vspace{0.30cm}

\normalsize
\textbf{This motivates interpolation methods such as Kriging.}


## What is Kriging?

\vspace{0.35cm}

\textbf{Definition}

\begin{itemize}
\item Kriging is a geostatistical interpolation method
\item Estimates values at unknown locations using nearby observed data
\item Uses both distance and spatial correlation between points
\item Provides prediction uncertainty
\end{itemize}

\vspace{0.25cm}

\begin{columns}[T]

\begin{column}{0.50\textwidth}

\textbf{Estimator}

\[
Z^*(x_0)=\sum_{i=1}^{n}\lambda_i Z(x_i)
\]

\begin{itemize}
\item $Z(x_i)$ = known observations
\item $\lambda_i$ = optimal weights
\item $Z^*(x_0)$ = predicted value
\item $n$ = number of points used
\end{itemize}

\end{column}

\begin{column}{0.50\textwidth}

\textbf{Core Idea}

\begin{itemize}
\item Closer points receive more weight
\item Farther points receive less weight
\item Weights are chosen optimally to improve prediction accuracy
\end{itemize}

\vspace{0.40cm}

\centering
\Large
Near $\Rightarrow$ More Influence \\[0.35cm]
Far $\Rightarrow$ Less Influence

\end{column}

\end{columns}

## Motivation for Kriging

\vspace{0.2cm}

\textbf{Why was Kriging developed?}

\begin{itemize}
\item Developed in the 1950s by South African mining engineer Danie Krige
\item Created to estimate gold grades from limited borehole samples/estimate ore concentrations in unsampled regions
\item Later formalized mathematically by Georges Matheron
\item Designed to improve prediction accuracy over simpler methods
\end{itemize}

\vspace{0.35cm}

\begin{columns}[T]

\begin{column}{0.48\textwidth}

\textbf{Why do we need it?}

\begin{itemize}
\item Real-world data is often sparse, time-consuming or expensive to collect
\item Measuring every location is impractical
\item Need reliable estimates between sampled points
\item Need uncertainty measures for decision-making
\end{itemize}

\end{column}

\begin{column}{0.48\textwidth}

\textbf{What is the goal?}

\begin{itemize}
\item Predict unknown values using nearby observations
\item Use spatial autocorrelation efficiently
\item Reduce bias in predictions
\item Minimize prediction error variance
\end{itemize}

\end{column}

\end{columns}

## Motivation for Kriging (cont.)

\vspace{0.3cm}

\textbf{When is Kriging a Good Choice?}

\begin{itemize}
\item Data has geographic or location-based patterns
\item Nearby points are likely related
\item Only limited observations are available
\item Accurate interpolation is needed between sampled locations
\item Prediction uncertainty is important for decisions
\end{itemize}

\vspace{0.55cm}

\centering
\Large
Spatial Data + Limited Samples $\Rightarrow$ Kriging

## Common Applications of Kriging

\vspace{0.35cm}

\begin{columns}[T]

\begin{column}{0.48\textwidth}

\textbf{Environmental \& Earth Sciences}

\begin{itemize}
\item Rainfall and temperature mapping
\item Air pollution estimation
\item Soil quality / agriculture mapping
\item Groundwater and hydrogeology
\item Oil, gas, and mineral exploration
\end{itemize}

\vspace{0.35cm}

\textbf{Modern Analytics}

\begin{itemize}
\item Public health risk mapping
\item Urban planning
\item Astronomy data interpolation
\end{itemize}

\end{column}

\begin{column}{0.48\textwidth}

\textbf{Engineering \& Industry}

\begin{itemize}
\item Real estate valuation
\item Circuit design optimization
\item Resource planning
\end{itemize}

\vspace{0.35cm}

\textbf{Remote Sensing}

\begin{itemize}
\item Filling missing spatial data
\item Surface prediction from satellite data
\end{itemize}

\vspace{0.35cm}

\textbf{Sports Analytics}

\begin{itemize}
\item Shot efficiency by location
\item Court / field heat maps
\item Player movement zones
\end{itemize}

\end{column}

\end{columns}

## Variations of Kriging

\vspace{0.25cm}

\textit{Several Kriging methods exist depending on assumptions about the mean and spatial trend.}

\vspace{0.35cm}

\begin{columns}[T]

\begin{column}{0.48\textwidth}

\textbf{Most Common Types}

\begin{itemize}
\item \textbf{Simple Kriging}  
Known constant mean

\item \textbf{Ordinary Kriging}  
Unknown but constant mean

\item \textbf{Universal Kriging}  
Allows a spatial trend in the mean/ Mean follows a trend across space or data
\end{itemize}

\end{column}

\begin{column}{0.48\textwidth}

\textbf{Other Extensions}

\begin{itemize}
\item Block Kriging
\item Co-Kriging
\item Indicator Kriging
\item Bayesian Kriging
\item Poisson Kriging
\end{itemize}

\end{column}

\end{columns}

\vspace{0.7cm}

\centering
\Large
Our Focus: \textbf{Ordinary Kriging}

## Advantages of Kriging

\vspace{0.3cm}

\begin{columns}[T]

\begin{column}{0.48\textwidth}

\textbf{Prediction Strength}

\begin{itemize}
\item Produces accurate estimates at unsampled locations
\item Uses spatial correlation, not only distance
\item Weights are chosen to minimize estimation variance
\item Often outperforms simple inverse distance weighting
\end{itemize}

\end{column}

\begin{column}{0.48\textwidth}

\textbf{Built-in Uncertainty}

\begin{itemize}
\item Predicts a value \textit{and} prediction error
\item Helps measure confidence in estimates
\item Useful when making real decisions from data
\item Identifies areas where more samples may be needed
\end{itemize}

\end{column}

\end{columns}

\vspace{0.45cm}

\textbf{Flexible Across Fields}

\begin{itemize}
\item Environment, health, geology, engineering, sports analytics
\end{itemize}

## Kriging - Overview

\begin{columns}
\begin{column}{0.55\textwidth}

\underline{Goal:} \\
- To estimate data values $Z^*(x_0)$ at unknown locations based on the given data $Z(x_i)$, for $i = 1, \dots, n$, where $n$ is the number of locations where data is observed. \\
- In mathematical terms,
\begin{align*}
Z^*(x_0) = \sum_{i=1}^n \lambda_i Z(x_i) = \mathbf{\lambda}^T \mathbf{z}
\end{align*}
\vspace{0.5em}

\underline{Idea:} \\
- An observation will be more similar to data that is "close" in space comparative to those "far" in space.
\vspace{1.5em}

"Closeness" will be determined by the spatial correlation between points in the given field of data.

\end{column}
\begin{column}{0.45\textwidth}

\includegraphics[width=\linewidth]{"dummy image - cmda4654 p2.png"}

\end{column}
\end{columns}





## The Kriging Model

\underline{Generic Model:}
\begin{align*}
Z(x) &= \mu(x) + \epsilon(x)
\end{align*}
$\epsilon(x)$ is assumed to have a zero mean and a corresponding variogram model with function $C(x_i, x_j)$. This is also notated as $C(\mathbf{h})$, where $\mathbf{h}$ is the difference vector between two locations $x_i$ and $x_j$. \
\vspace{1em}
Note: This assumes that the variation in the data is a function of the distance between points.
\vspace{1em}

There are a handful of different types of Kriging but we will be focusing on *Ordinary Kriging*. \
\vspace{1.5em}
Ordinary Kriging assumes an unknown stationary mean. That is, the mean of the response variable is independent of location. This simplifies the above model to:
\begin{align*}
Z(x) = \mu + \epsilon(x)
\end{align*}
Additionally, we assume isotropy; any pair of locations that are the same distance from each other, regardless of direction, have the same spatial correlation.

## Solving for the Kriging Weights

To solve for the weights, $\lambda_i$ in
$$Z^*(x_0) = \sum_{i=1}^n \lambda_i Z(x_i),$$
To estimate the weights $lambda_i$ is equivalent to finding the $lambda_i$ that minimizes the error variance  
$$\text{min} \ \sigma^2_e = \text{min }\text{Var}\left[ Z(x) - Z^*(x) \right]$$
such that the estimator is unbiased, i.e. $\text{E}\big[Z(x) - Z^*(x)\big] = 0$. (This is satisfied by assuming a stationary mean.) \
So, the weights are constrained to satisfy
$$\sum_{i=1} ^n \lambda_i = 1  \hspace{1em}  \text{ or } \hspace{1em} \mathbf{\lambda}^T \mathbf{1} = 1$$
and we are able to express the minimization problem as,
$$\text{min} \ \sigma^2_e = \text{min }E \big[ (Z(x) - Z^*(x))^2 \big]$$

To solve this constrained optimization problem, the method of Lagrange multiplier is used.

## Solving for the Kriging Weights (cont'd)

Not all proof steps are shown, but a thorough reference \href{http://www.stat.ucla.edu/~nchristo/statistics_c173_c273/c173c273_ordinary_kriging_using_covariance.pdf}{here} provides a clear explanation. \
\vspace{0.5em}
After some algebraic manipulation, we arrive at minimizing
\begin{align*}
&C(0) - 2 \sum_{i=1}^n \lambda_i C(x_0, x_i) + \sum_{i=1}^n \sum_{j=1}^n \lambda_i \lambda_j C(x_i, x_j) -2 \Lambda \left( \sum_{i=1}^n \lambda_i - 1 \right)\\
&= \mathbf{c_0} - 2 \mathbf{\lambda}^T \mathbf{c_{x_0}} + \mathbf{\lambda}^T \mathbf{C} \mathbf{\lambda} - 2 \Lambda \left( \mathbf{\lambda}^T \mathbf{1} - \mathbf{1} \right)
\end{align*}
by taking the partial derivatives with respect to the $\lambda_i$'s (or vector $\mathbf{\lambda}$) and then the lagrange multiplier $\Lambda$. \
\vspace{0.5em}
The solution results in solving the system
\begin{align*}
\begin{bmatrix}
C(x_1, x_1) & C(x_1, x_2) & C(x_1, x_3)  & \cdots & C(x_1, x_n) & 1 \\
C(x_2, x_1) & C(x_2, x_2) & C(x_2, x_3)  & \cdots & C(x_2, x_n) & 1 \\
C(x_3, x_1) & C(x_3, x_2) & C(x_3, x_3)  & \cdots & C(x_3, x_n) & 1 \\
\vdots &      \vdots &         & \ddots           &         & 1 \\
C(x_n, x_1) & C(x_n, x_2) & C(x_n, x_3)  & \cdots & C(x_n, x_n) & 1 \\
1 & 1 & 1  & \cdots & 1 & 0
\end{bmatrix}
\begin{bmatrix}
\lambda_1 \\ \lambda_2 \\ \lambda_3 \\ \vdots \\ \lambda_n \\ \Lambda   
\end{bmatrix} 
 = 
 \begin{bmatrix}
 C(x_0, x_1) \\  C(x_0, x_2) \\  C(x_0, x_3) \\ \vdots \\  C(x_0, x_n) \\ 1 
 \end{bmatrix} \hspace{1em}
\end{align*}



## Solving for the Kriging Weights (cont'd)

The system given on the previous slide can be equivalently expressed as
\begin{align*}
\begin{bmatrix} \mathbf{C} & \mathbf{1} \\ \mathbf{1}^T & 0 \end{bmatrix} \begin{bmatrix} \mathbf{\lambda} \\ \Lambda \end{bmatrix} = \begin{bmatrix} \mathbf{c}_0 \\ 0 \end{bmatrix}
\end{align*}
And solved by simply inverting the matrix on the left hand side. \
\vspace{0.5em}
This matrix will be *in theory* invertible. The function $C(x_i, x_j)$ arises from variogram modeling, which uses conditionally negative definite functions. Mathematical depth aside, this class of functions will produce a positive definite matrix (which is invertible). \
Note: As $n$ becomes large, the dimensions of the matrix could cause it to become ill-conditioned.\
\vspace{0.5em}
After solving for the kriging weights, we can find the error variance of our estimate $Z^*(x_0)$ as
\begin{align*}
\sigma^2_e = C(0) + \sum_{i=1}^n \lambda_i C(x_0, x_i) + \Lambda
\end{align*}

## Variogram Model

We previously assumed that the spatial covariance can be modeled with a variogram function $2\gamma(\mathbf{h})$, or equivalently a semivariogram function $\gamma(\mathbf{h})$. (Though in spatial statistics these two terms are sometimes used interchangeably.)  \
In either case, we seek a function that describes the spatial covariance for the distance between two points. \

This is achieved by first obtaining a *sample variogram* that is described by
\begin{align*}
2 \widehat{\gamma} (\mathbf{h}) = \frac{1}{\left| N_{ \mathbf{h}}\right|} \sum_{N_{\mathbf{h}}} \left( Z(x_i) - Z(x_j) \right)^2
\end{align*}
where $\left| N_{ \mathbf{h}}\right|$ is the number of unique pairs in the set defined as
\begin{align*}
N_{ \mathbf{h}} := \left\{ (x_i, x_j) : | x_i - x_j| = h \right\} \text{ for } i, j = 1, \dots, n
\end{align*}

That is, we obtain $k$ discrete set of points with distances $h$ and corresponding average of variances $2\widehat{\gamma}(\mathbf{h})$.


## Variogram Model (cont'd)

Last, a variogram model is fit to the sample variogram. The optimal variogram will minimize the sum of squared error given by
\begin{align*}
\sum_{i=1}^k \big( 2 \widehat{\gamma}(\mathbf{h}) - 2 \gamma(\mathbf{h}, \alpha )\big)^2
\end{align*}
Where $\alpha$ is used to denote parameters of the variogram model. \
\vspace{1.5em}
For example one type of model is a spherical semivariogram. Here, the parameters are the "nugget", "partial sill" and "range" in geospatial statistic terms. Equivalently, these dictate the y-intercept, the difference in the maximum covariance and the y-intercept and the distance between the origin and where the maximum covariance is reached. \
\vspace{1em}
Once a model has been fit, this function can be used to determine the values of $C(x_i, x_j)$ in the matrix $\begin{bmatrix} \mathbf{C} & \mathbf{1} \\ \mathbf{1}^T & 0 \end{bmatrix}$ and that of $C(x_0, x_i)$ in the vector $\begin{bmatrix} \mathbf{c}_0 \\ 0 \end{bmatrix}$.

## Basic Example

I have created a basic and small data set to demonstrate how to compute
Kriging. The x and y variables represent testing locations within a lake, and
the z variable is the concentration of algae at each testing location. The
pred value is a coordinate without a testing area that we want to estimate
algae concentration at.

```{r algae, echo = TRUE}
set.seed(1000)
dat = data.frame(x = rnorm(6, mean=5, sd=3), y = rnorm(6, mean=5, sd=3),
                     z = rnorm(6, mean=0.5, sd=0.3))
pred = data.frame(x = 2, y = 3)
```

## Semivariance and Distance

Here I calculate the distance between each unique set of points and the semivariance
for each set of points.

```{r dist_semivariance, echo = TRUE}
semivariogram = data.frame(matrix(ncol = 2, nrow = 0))
colnames(semivariogram) = c('Distance', 'Semivariance')
for(i in 1:nrow(dat)){
  for(j in 1:nrow(dat)){
    if(j > i){
      dist = sqrt((dat$x[i] - dat$x[j])^2 + (dat$y[i] - dat$y[j])^2)
      semivariance = 0.5*((dat$z[i] - dat$z[j])^2)
      semivariogram = rbind(semivariogram, data.frame("Distance" = dist, 
                                                      "Semivariance" = semivariance))
    }
  }
}
```

## Semivariogram Plot

```{r semivariogram, echo = TRUE, fig.width=4, fig.height=2.33}
library(ggplot2)
ggplot(semivariogram, aes(x = Distance, y = Semivariance)) + geom_point() +
  ggtitle("Semivariogram for the Lake Data")
```

## Semivariogram Model Selection

There are a few different widely used semivariogram models. The most simple
model is a linear model, and for this model there is no point in the graph
where the semivariance flattens out and stops increasing, so the sill and 
range must be arbitrarily selected. Another common model is the spherical
model which appears as a quadratic function that flattens out where the
semivariance begins to flatten out. The exponential model is similar to the
spherical model but does not flatten out as completely.

We determine the nugget, sill, and range for the model. There is no y-intercept
in the semivariogram plot, so we can assume that the model intercepts the y-axis
at the origin.

```{r semivariance_parameters, echo = TRUE}
nugget = 0
partial_sill = max(semivariogram$Semivariance) - nugget
range = semivariogram$Distance[10]
```

An exponential model can be used for the semivariogram of this data, and this
will be applied to a distance matrix of the points.

```{r exponential_model, echo = TRUE}
dist_mat = as.matrix(dist(dat[, 1:2]))
semivariogram2 = nugget + partial_sill*exp(-dist_mat/range)
```

## Set up Matrices and Vectors

I also calculate the vector for the Euclidean distance between the points in the plane
and the new testing coordinates.

```{r pred_vec, echo = TRUE}
dist_pred = sqrt((dat$x - pred$x)^2 + (dat$y - pred$y)^2)
```

I set up the matrices by adding a column and row of ones with a 0 in the bottom
right spot in the matrix. This allows us to solve for the lambda and lagrange
multiplier values.

```{r matrices, echo = TRUE}
semi_mat = matrix(1, 7, 7)
semi_mat[1:6, 1:6] = semivariogram2
semi_mat[7, 7] = 0
```

I calculate the semivariogram for the vector of the distances between the points
and the new prediction coordinates.

```{r pred_semivariance, echo = TRUE}
semi_dist = nugget + partial_sill*exp(-dist_pred/range)
dist_pred2 = c(semi_dist, 1)
```

## Solving for Weights

We can solve for the weights and the lagrange multiplier, v, by taking
the inverse of the matrix and multiplying it by the vector for the distances between
the points and the new prediction coordinates.

```{r weights, echo = TRUE}
weights_v = solve(semi_mat) %*% dist_pred2
```

The weights are the first n values of the resulting vector, where n is the number
of rows in the original data frame. The lagrange multiplier is the final value in
the vector, and this is used to calculate the variance of the prediction.

```{r weights_lagrange, echo = TRUE}
weights = weights_v[1:6]
lagrange = weights_v[7]
```

## Prediction

The predicted value is the sum of each weight multiplied by the corresponding
z value for the coordinates from the original data frame.

```{r prediction, echo = TRUE}
z_pred = sum(weights*dat$z)
```

The variance of the errors is the dot produce of the transpose of the weight 
vector and the semivariogram vector of the distances between the original 
points and the new coordinates plus the lagrange multiplier value.

```{r variance_of_errors, echo = TRUE}
variance = t(weights) %*% semi_dist + lagrange
```

## Plot of Prediction

```{r combine_pred_and_dat, echo = FALSE}
pred2 = c(2, 3, z_pred)
dat2 = rbind(dat, pred2)
```

```{r prediction_plot, echo = TRUE, fig.width=4, fig.height=2.5}
ggplot(dat2, aes(x = x, y = y, size = z)) + 
  geom_point() + 
  ggtitle("Locations sized by Algae Concentration")
```

## Quakes - Introduction

We are going to look at an example using the quakes dataset which is built into r. This dataset has 1000 observations of earthquake data around the Fiji area and includes things like the earthquakes location via latitude and longitude, the depth of the earthquake in km, the earthquakes magnitude, and how many seismic stations detected the earthquake. This example is going to look into what kind of insights and predictions can be made on the magnitude of earthquakes through the use of kriging.

$~$

```{r, echo=TRUE}
library(sf)
library(gstat)
library(mapview)
library(stars)

data(quakes)

# Convert data into a spatial object
quakes_sf <- st_as_sf(quakes, coords = c("long", "lat"), crs = 4326) %>%
  st_transform(32760)
```
```{r, echo=TRUE, eval=FALSE}

# Plot on map
mapview(quakes_sf, 
        zcol = "mag",
        map.types = "CartoDB.Voyager", 
        legend = TRUE)
```

## Quakes - Plotted data

\includegraphics[width=\linewidth]{"plotted data.png"}

## Quakes - Variogram

The Variogram is one of the most important and insightful parts of kriging. The variogram is first estimated from the data, and then we choose a theoretical model to be fitted to it. There is no "best" method when it comes to choosing a model, but the variogram helps guide you to choosing the best fit for your data. Choosing the best possible model that accurately represents the spatial structure of your data is very important, as it directly impacts the quality of your predictions.

$~$

```{r, echo=TRUE, eval=FALSE}
# Variogram
v <- variogram(mag ~ 1, quakes_sf)
# Selected Model
v_fit <- fit.variogram(v, vgm(psill = 0.2, model = "Sph", range = 150000, nugget = 0.05))
# Plotting
v_df <- as.data.frame(v)
v_line <- variogramLine(v_fit, maxdist = max(v_df$dist))
v_line_df <- as.data.frame(v_line)

ggplot() +
  geom_point(data = v_df, aes(x = dist, y = gamma)) +
  geom_line(data = v_line_df, aes(x = dist, y = gamma)) +
  labs(
    title = "Variogram with Fitted Model",
    x = "Distance",
    y = "Semivariance"
  ) +
  theme_minimal()
```

## Quakes - Variogram
```{r, fig.width=5.5, fig.height=3, warning=FALSE, results='hide'}
# Variogram
v <- variogram(mag ~ 1, quakes_sf)
# Model
v_fit <- fit.variogram(v, vgm(psill = 0.2, model = "Sph", range = 150000, nugget = 0.05))
v_df <- as.data.frame(v)
# Plotting
v_line <- variogramLine(v_fit, maxdist = max(v_df$dist))
v_line_df <- as.data.frame(v_line)

ggplot() +
  geom_point(data = v_df, aes(x = dist, y = gamma)) +
  geom_line(data = v_line_df, aes(x = dist, y = gamma)) +
  labs(
    title = "Variogram with Fitted Model",
    x = "Distance",
    y = "Semivariance"
  ) +
  theme_minimal()
```

## Quakes - Kriging

Now that we have our model, we need some new points that we want to predict the magnitude for using kriging. To do this, we are going to create a grid of points and project it onto the Fiji area. This will allow us to use the points we have data for to make predictions on every point in the grid to see what their potential magnitude could look like. 

Putting all this together, we can use the krige function from the gstat library to make predictions on the grid we just made and plot it onto a map.

$~$

```{r, echo=TRUE}
# create predictions grid in the Fiji area
pred_grid <- st_bbox(c(xmin = 163, ymin = -30, xmax = 185, ymax = -8), crs = 4326)
quakes_grid <- st_as_stars(pred_grid, dx = 0.5, dy = 0.5) %>%
  st_as_sf() %>%
  st_transform(32760)
```
```{r, echo=TRUE, eval=FALSE}

# Kriging
quakes_kriged <- krige(mag ~ 1, quakes_sf, quakes_grid, model = v_fit)
quakes_kriged_map <- st_transform(quakes_kriged, 4326)
quakes_fin <- st_shift_longitude(quakes_kriged_map)
names(quakes_fin)[3] <- "pred mag"

# Plot on map
mapview(quakes_kriged, 
        zcol = "pred mag")
```

## Quakes - Predictions

\begin{figure}[h]
\centering
\includegraphics[width=0.795\linewidth]{"mag prediction grid.png"}
\end{figure}

## Quakes - Limitations and Conclusion

**Limitations**

\begin{itemize}
  \item Large areas with few data points tend to get pulled towards the global mean
  \item Extreme values have been "weakened" and pulled closer to the global mean
  \item Earthquakes have high variability when it comes to their magnitude
\end{itemize}

$~$

**Conclusion**

Our results give a decent idea of the general magnitude of earthquakes in areas we have data points for, but isn't great for predicting the exact magnitude and it also struggles with predicting the magnitude in areas that lack data points. This highlights the idea that kriging is suited more towards interpolation rather than extrapolation. Kriging prefers smaller and more tight data, so trying make predictions over a large area is going to result in over smoothing. This could be improved in this example by taking a smaller grid in a specific area of interest or by making the grid a more unique shape that follows the shape of the data rather than trying to get estimates for areas where we have no data points.


```{r load-nba-data, cache=TRUE, include=FALSE}
# Read in NBA play-by-play data for the 2025-26 season
pbp <- load_nba_pbp(seasons = 2026)
player_box <- load_nba_player_box(seasons = 2026)

# Read in a dictionary of player names and their IDs for the NBA data
player_names <- player_box %>% distinct(athlete_id, athlete_display_name)
```

```{r, include=FALSE}
# Unprinted Code for NBA Example

# Constants
rim_x <- 5.25
three_pt_r <- 23.75
corner_x <- rim_x + sqrt(three_pt_r^2 - 22^2)
paint_x <- 19
paint_y <- 8
bin_size <- 2
min_attempts <- 10

# Function to draft the lines of a basketball halfcourt
draw_court <- function() {
  
  # Define the angle of the three point line
  angle_top <- atan2(22, corner_x - rim_x)
  angle_bot <- atan2(-22, corner_x - rim_x)
  
  # Build the three point line using arc points
  arc_points <- data.frame(
    x = rim_x + three_pt_r * cos(seq(angle_top, angle_bot, length.out = 1000)),
    y = three_pt_r * sin(seq(angle_top, angle_bot, length.out = 1000))
  )
  
  # Make a list of the output elements
  list(
    # Three point arc
    geom_path(data = arc_points, aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # Corner three lines
    geom_path(data = data.frame(x = c(0, corner_x), y = c(22, 22)),
              aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    geom_path(data = data.frame(x = c(0, corner_x), y = c(-22, -22)),
              aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # Paint
    geom_path(data = data.frame(x = c(0, 19, 19, 0, 0), y = c(8, 8, -8, -8, 8)),
              aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # FT circle solid
    geom_path(data = data.frame( x = 19 + 6 * cos(seq(-pi/2, pi/2, length.out = 500)),
                                 y = 6 * sin(seq(-pi/2, pi/2, length.out = 500))), aes(x = x, y = y),
              inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # FT circle dashed
    geom_path(data = data.frame(x = 19 + 6 * cos(seq(pi/2, 3*pi/2, length.out = 500)),
                                y = 6 * sin(seq(pi/2, 3*pi/2, length.out = 500))), aes(x = x, y = y),
              inherit.aes = FALSE, color = "black", linewidth = 0.6, linetype = "dashed"),
    # Restricted arc
    geom_path(data = data.frame(x = rim_x + 4 * cos(seq(-pi/2, pi/2, length.out = 500)),
                                y = 4 * sin(seq(-pi/2, pi/2, length.out = 500))), aes(x = x, y = y),
              inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # Rim
    geom_path(data = data.frame(x = rim_x + 0.75 * cos(seq(0, 2*pi, length.out = 500)),
                                y = 0.75 * sin(seq(0, 2*pi, length.out = 500))), aes(x = x, y = y),
              inherit.aes = FALSE, color = "darkorange2", linewidth = 0.8))
}

# Data Cleaning
shots <- pbp %>%
  
  # Filter to plays that have shooting coordinates
  filter(shooting_play == TRUE,
         !is.na(coordinate_x),
         !is.na(coordinate_y),
         # Remove free throws
         !grepl("Free Throw", type_text),
         #Filter to the regular season
         season_type == 2) %>%
    
  # Translate from player IDs to names
  left_join(player_names, by = c("athlete_id_1" = "athlete_id")) %>%
    
  #Flip the data into being only one half of the court
  mutate(team_name = ifelse(team_id == home_team_id,
         paste(home_team_name, home_team_mascot), paste(away_team_name, away_team_mascot)),
    x = ifelse(coordinate_x > 0, -(coordinate_x - 46.75), coordinate_x + 46.75),
    y = ifelse(coordinate_x > 0, -coordinate_y, coordinate_y)) %>%
  
  # Get rid of heaves from beyond half court
  filter(x <= 45) %>%
    
  # Narrow the columns again
  select(athlete_display_name, team_name, score_value, x, y)

  
# Helper function to make the prediction grid
make_pred_grid <- function() {
  
  # Make a grid of squares across the court
  pred_grid <- expand.grid(
    x = seq(0.5, 45.5, by = 1),
    y = seq(-24.5, 25.5, by = 1)) %>%
    as.data.frame() %>%
    
    # Calculate the distance from the rim of each point
    mutate(dist_from_rim = sqrt((x - rim_x)^2 + y^2)) %>%
    
    # Filter to points that are within 5 feet of the three point line
    filter(dist_from_rim <= three_pt_r + 5) %>%
    select(x, y) %>%
    as.data.frame()
  
  # Make the grid itself
  coordinates(pred_grid) <- ~x+y
  gridded(pred_grid) <- TRUE
  pred_grid
}

# Helper function to make the bins for a team
make_bins <- function(team, bin_size = 2, min_attempts = 10) {
  
  # Filter to only shot attempts from a specified team
  team_shots %>%
    filter(team_name == team) %>%
    
    # Create 2 ft x 2 ft bins across the court
    mutate(
      x_bin = floor(x / bin_size) * bin_size + bin_size / 2,
      y_bin = floor(y / bin_size) * bin_size + bin_size / 2) %>%
    group_by(x_bin, y_bin) %>%
    
    # Define the attempts, makes, and field goal percent of each bin
    summarise(
      attempts = n(),
      makes = sum(score_value > 0),
      fg_pct = makes / attempts,
      .groups = "drop") %>%
    
    # Ensure the bin has enough attempts to be meaningful
    filter(attempts >= min_attempts) %>%
    
    # Add characteristics of each bin like point value
    mutate(
      dist_from_rim = sqrt((x_bin - rim_x)^2 + y_bin^2),
      is_three = dist_from_rim >= three_pt_r | (abs(y_bin) >= 22 & x_bin <= corner_x),
      in_paint = x_bin <= paint_x & abs(y_bin) <= paint_y,
      zone = case_when(
        is_three ~ "Three",
        in_paint ~ "Paint",
        TRUE     ~ "Midrange")) %>%
    as.data.frame()
}
```

## Data Introduction

Points per shot is one of the biggest current topics of discussion in basketball analytics. It is calculated using two factors: a team or players chance of making a shot from a position on the floor and the point of value of a make from that position. Performing analysis on this topic requires gathering both of those values which can be done using the `hoopr` package.

```{r, echo=TRUE, eval=FALSE}
library(hoopR)

pbp <- load_nba_pbp(seasons = 2026)
player_box <- load_nba_player_box(seasons = 2026)

player_names <- player_box %>%
  distinct(athlete_id, athlete_display_name)

shots <- pbp %>%
  filter(shooting_play == TRUE,
         !is.na(coordinate_x), !is.na(coordinate_y), !grepl("Free Throw", type_text),
         season_type == 2) %>%
  select(athlete_id_1, coordinate_x, coordinate_y, score_value) %>%
  left_join(player_names, by = c("athlete_id_1" = "athlete_id")) %>%
  mutate(
    x = ifelse(coordinate_x > 0,
               -(coordinate_x - 46.75),
               coordinate_x + 46.75),
    y = ifelse(coordinate_x > 0, -coordinate_y, coordinate_y)) %>%
  filter(x <= 45) %>%
  select(athlete_display_name, score_value, x, y)
```

## Data Example

The data has been loaded and saved as csv for ease of access: \href{https://docs.google.com/spreadsheets/d/1uBDuThoKEiOK03TE7cipW_oFBo6iOxBon5RDeMG9lUI/edit?usp=sharing}{Click here} to access shots.csv. 
\vspace{0.15cm}
\begin{columns}
\begin{column}{0.5\textwidth}

Preview of the data
\vspace{0.1cm}
```{r}
# Look at the first 20 rows of the shots data
shots %>% 
  head(20) %>%
  knitr::kable(format = "latex", booktabs = TRUE)
```

\end{column}
\begin{column}{0.5\textwidth}

\quad{}\quad{}\quad{}NBA 2025-26 Regular Season Heatmap
\vspace{0.1cm}
```{r, fig.width=4, fig.height=2.33}
# Make a plot of the NBA heatmap from the 2025-26 season
ggplot(shots, aes(x = x, y = y)) +
  stat_density_2d(aes(fill = after_stat(density)),geom = "raster",contour = FALSE,alpha = 0.9) +
  
  # Draw the court lines
  draw_court() +
  scale_fill_viridis_c() +
  
  #Establish the boundaries
  coord_fixed(xlim = c(0, 45),ylim = c(-25, 25),expand = FALSE,clip = "off") +
  
  # Add axis labels
  labs(title = "NBA Shot Locations",x = "Distance from Basket (ft)",y = "Court Width (ft)",fill = "Density") +
  
  # Adjust the plot positioning
  theme_minimal(base_size = 7) + 
  theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = -50))
```
\end{column}
\end{columns}

## Kriging Motivation

It is clear from the shot density layout on the previous slide that there are two main areas that teams take shots from:

\begin{itemize}
  \item the three point line
  \item the rim
\end{itemize}

This makes sense given that shots at the rim are likely to be high percentage and shots from three point range are worth more points. The middle area between these two hot zones is fairly empty, and the league has been trending to make midrange jumpshots even less a part of offenses in the Modern Era. All of this evolution is in the pursuit of offensive efficiency which is where Kriging comes in.

Since the shot map is made up of x and y data, it can be treated similarly to a geostatistics field from more typical Kriging -  it has high density areas and low gaps with a large number of very individualized spots. Kriging can be used to smooth over these less commonly attempted areas and give a picture of not only where teams shoot but how well they perform from these areas.

Combining those two areas will mean that Kriging can make a points per shot estimation from every location on the floor which can then be used to check if teams are running their offenses to peak efficiency.

## Attempt Frequency Kriging

A typical picture of attempt frequency for a basketball team would consist of colored boxes or hexes that represent volume from each area of the court, but a custom Kriging can smooth this out into a continuous map using two foot by two foot bins.

```{r plot-attempts}
# Get the shots for a specific team
team_shots <- shots %>% select(team_name, score_value, x, y)

# Define a function to perform Kriging on the shot attempts of a team
plot_attempts <- function(team) {
  
  # Use helper functions to build correct bins
  pred_grid <- make_pred_grid()
  bins <- make_bins(team)
  attempt_bins <- bins %>% select(x_bin, y_bin, attempts) %>% as.data.frame()
  coordinates(attempt_bins) <- ~x_bin+y_bin
  
  # Run autoKrige on the shot attempts data
  krig_attempts <- suppressWarnings(autoKrige(attempts ~ 1, input_data = attempt_bins, new_data = pred_grid))
  
  # Form a dataframe from the Kriging output
  krig_attempts_df <- as.data.frame(krig_attempts$krige_output)
  names(krig_attempts_df) <- c("x", "y", "attempts_pred", "attempts_var", "attempts_stdev")
  krig_attempts_df <- krig_attempts_df %>% mutate(attempts_pred = pmax(0, attempts_pred))
  
  # Plot the output using helper functions
  ggplot(krig_attempts_df, aes(x = x, y = y, fill = attempts_pred)) +
    geom_raster(interpolate = TRUE) +
    draw_court() +
    scale_fill_viridis_c() +
    coord_fixed(xlim = c(0, 35), ylim = c(-25, 25)) +
    labs(title = paste(team, "- Shot Attempts"),
         x = "Distance from Basket (ft)",
         y = "Court Width (ft)",
         fill = "Attempts") +
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = -3))
}
```
\begin{columns}
\begin{column}{0.5\textwidth}

```{r bos-attempts-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_attempts("Boston Celtics")
```

\end{column}
\begin{column}{0.5\textwidth}

```{r det-attempts-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_attempts("Detroit Pistons")
```

\end{column}
\end{columns}

## FG\% Kriging

While the Kriging for shot attempts is somewhat trivial, it is extremely useful for field goal percentage. This statistic leads to a map that has continuous values across all the bins and the maps will differ more greatly team-to-team.

```{r plot-percentage}
# Define a function to perform Kriging on the field goal percentage of a team
plot_fg_pct <- function(team) {
  
  # Call helper functions to build the correct bins
  pred_grid <- make_pred_grid()
  bins <- make_bins(team)
  fg_bins <- bins %>% select(x_bin, y_bin, fg_pct) %>% as.data.frame()
  coordinates(fg_bins) <- ~x_bin+y_bin
  
  # Call autoKrige on the field goal percentage data
  krig_fg_auto <- suppressMessages(suppressWarnings(autoKrige(fg_pct ~ 1, input_data = fg_bins, new_data = pred_grid)))
  
  # Create a df from the Kriging output
  krig_fg_df <- as.data.frame(krig_fg_auto$krige_output)
  names(krig_fg_df) <- c("x", "y", "fg_pred", "fg_var", "fg_stdev")
  
  # Return both a plot of the output and the data itself
  return(list(
    plot = ggplot(krig_fg_df, aes(x = x, y = y, fill = fg_pred)) +
      geom_raster(interpolate = TRUE) +
      draw_court() +
      scale_fill_viridis_c(labels = scales::percent,
                           limits = c(0.20, 0.60),
                           oob = scales::squish) +
      coord_fixed(xlim = c(0, 35), ylim = c(-25, 25)) +
      labs(title = paste(team, "- FG%"),
           x = "Distance from Basket (ft)",
           y = "Court Width (ft)",
           fill = "FG%") +
      theme_minimal(base_size = 7) +
      theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = -3)),
    data = krig_fg_df,
    krig = krig_fg_auto))
}
```
\begin{columns}
\begin{column}{0.5\textwidth}

```{r bos-percentage-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
bos_fg <- plot_fg_pct("Boston Celtics")
bos_fg$plot
```

\end{column}
\begin{column}{0.5\textwidth}

```{r det-percentage-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
det_fg <- plot_fg_pct("Detroit Pistons")
det_fg$plot
```

\end{column}
\end{columns}

## Points Per Shot

As mentioned previously, points per shot is a metric that is used to determine the quality of a type of shot attempt. Now that the field goal percentage Kriging has supplied a likelihood of making a shot at any location on the floor, it is easy to take it a step further into point values of the shots.

```{r}
# Define a function to perform Kriging on the points per shot data
plot_pps <- function(fg_data, team = "") {
  
  # Define the number of points each shot is worth
  fg_data %>%
    mutate(dist_from_rim = sqrt((x - rim_x)^2 + y^2),
      is_three = dist_from_rim >= three_pt_r | (abs(y) >= 22 & x <= corner_x),
      point_value = ifelse(is_three, 3, 2),pps = fg_pred * point_value) %>%
    
    # Plot the data using helper functions
    ggplot(aes(x = x, y = y, fill = pps)) +
    geom_raster(interpolate = TRUE) +
    draw_court() +
    scale_fill_viridis_c(limits = c(0.5, 2.0), oob = scales::squish) +
    coord_fixed(xlim = c(0, 35), ylim = c(-25, 25)) +
    labs(title = paste(team, "- Points per Shot"),
         x = "Distance from Basket (ft)",
         y = "Court Width (ft)",
         fill = "PPS") +
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = -3))
}
```
\begin{columns}
\begin{column}{0.5\textwidth}

```{r bos-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(bos_fg$data, "Boston Celtics")
```

\end{column}
\begin{column}{0.5\textwidth}

```{r det-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(det_fg$data, "Detroit Pistons")
```

\end{column}
\end{columns}

## Non-Kriging Approach

So the question becomes "How do these smooth maps help NBA teams improve their offenses?"

One of the main advantages that these plots provide is the way they make it much easier to full picture of slots across the floor. As a comparison, here would be a more standard look at a teams percentages and shot volumes from Synergy Sports. 

\begin{columns}
\begin{column}{0.5\textwidth}
The plot makes it hard to tell whether there's smaller deviations within the large group and doesn't calculate the point values of the shots (i.e. shooting 50\% from 2 is the same as 33\% from 3). Thus it is generally not very insightful about which shots are the most efficient, only offering general statistics about which areas the team has shot the most accurately from.
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=\linewidth]{"Synergy Shot Chart.png"}
\end{column}
\end{columns}

## Kriging Comparison

A smooth chart from Kriging much better captures the specific peaks and valleys of the shooting percentage and factoring in the point values gives real, actionable insights. Combining this with the shots attempted map can make it easy for a coach without a statistical background to see mismatches in shot value and frequency without complex interpretation.

Running through every NBA team can make some interesting comparisons:

```{r, eval=FALSE, echo=TRUE}
plot_team_impact <- function(team_name) {
   fg_obj <- plot_fg_pct(team_name)
   attempts_plot <- plot_attempts(team_name)
   pps_plot <- plot_pps(fg_obj$data)
   attempts_plot | pps_plot
}

nba_teams <- sort(unique(team_shots$team_name))

for (tm in nba_teams) {
   print(plot_team_impact(tm))
}
```

## Team Takeaways

```{r, fig.width=3, fig.height=3, warning=FALSE, results='hide'}
# Interesting comparisons
cle <- plot_fg_pct("Cleveland Cavaliers")$data


# Interesting side-to-side plots
ind <- plot_fg_pct("Indiana Pacers")$data
den <- plot_fg_pct("Denver Nuggets")$data
```

Some teams could gain insights by looking at their charts side-by-side, like the Cleveland Cavaliers

\begin{columns}
\begin{column}{0.5\textwidth}
```{r cle-attempts-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_attempts("Cleveland Cavaliers")
```
\end{column}
\begin{column}{0.5\textwidth}
```{r cle-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(cle, "Cleveland Cavaliers")
```
\end{column}
\end{columns}

## Points per Shot Patterns

Some teams also have clear patterns that favor shooting from one side of the floor

\begin{columns}
\begin{column}{0.5\textwidth}
```{r ind-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(ind, "Indiana Pacers")
```
\end{column}
\begin{column}{0.5\textwidth}
```{r den-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(den, "Denver Nuggets")
```
\end{column}
\end{columns}

## Accuracy of Predictions

Of course, the predicted maps are only useful if they are accurate. So how well did Kriging do with smoothing the shooting data?

```{r}
# Define a function to show the accuracy of the Kriging predictions
plot_accuracy <- function(team) {
  
  # Call the helper functions
  bins <- make_bins(team)
  krig_fg_df <- plot_fg_pct(team)$data
  
  # Adjust the bins to match the real data
  bins %>%
    mutate(x_snap = round((x_bin - 0.5)) + 0.5, y_snap = round((y_bin - 0.5)) + 0.5) %>%
    left_join(krig_fg_df, by = c("x_snap" = "x", "y_snap" = "y")) %>%
    filter(!is.na(fg_pred)) %>%
    
    # Make a plot of the accuracy
    ggplot(aes(x = fg_pct, y = fg_pred, color = zone, size = attempts)) +
    geom_point(alpha = 0.75) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.6) +
    
    # Color code the plot by shot type
    scale_color_manual(values = c("Paint" = "#C0392B", "Midrange" = "#F39C12", "Three" = "#2471A3")) +
    scale_size_continuous(range = c(1, 5), guide = "none") +
    scale_x_continuous(labels = scales::percent, limits = c(0.15, 0.85)) +
    scale_y_continuous(labels = scales::percent, limits = c(0.15, 0.85)) +
    
    # Add labels
    labs(title = paste(team, "- Kriging Accuracy"), x = "Actual FG%", y = "Kriging Predicted FG%", color = "Zone") +
    
    # Adjust the positioning of the plot
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = 2))
}
```

\begin{columns}
\begin{column}{0.5\textwidth}
```{r bos-acc-plot, fig.width=2.75, fig.height=2.4, warning=FALSE, results='hide'}
plot_accuracy("Boston Celtics")
```
\end{column}
\begin{column}{0.5\textwidth}
```{r det-acc-plot, fig.width=2.75, fig.height=2.4, warning=FALSE, results='hide'}
plot_accuracy("Detroit Pistons")
```
\end{column}
\end{columns}

## Kriging Limitations

```{r}
# Define a function to plot the field goal percentage Kriging of a specific player
plot_fg_pct_player <- function(player) {
  
  # Define a fully expanded half court grid for the players
  pred_grid <- expand.grid(x = seq(0.5, 45.5, by = 1),y = seq(-24.5, 25.5, by = 1)) %>%
    as.data.frame() %>% select(x, y) %>% as.data.frame()
  coordinates(pred_grid) <- ~x+y
  gridded(pred_grid) <- TRUE
  
  # Re-define the bins to match this new grid
  fg_bins <- shots %>%
    filter(athlete_display_name == player) %>%
    mutate(x_bin = floor(x / bin_size) * bin_size + bin_size / 2,
      y_bin = floor(y / bin_size) * bin_size + bin_size / 2) %>%
    group_by(x_bin, y_bin) %>%
    summarise(
      attempts = n(),
      makes = sum(score_value > 0),
      fg_pct = makes / attempts,
      .groups = "drop") %>%
    filter(attempts >= 3) %>%
    as.data.frame()
  coordinates(fg_bins) <- ~x_bin+y_bin
  
  # Run Kriging on the individual field goal percentage data
  krig_fg <- suppressMessages(suppressWarnings(autoKrige(fg_pct ~ 1, input_data = fg_bins, new_data = pred_grid)))
  
  # Form a data frame on the Kriging output
  krig_fg_df <- as.data.frame(krig_fg$krige_output)
  names(krig_fg_df) <- c("x", "y", "fg_pred", "fg_var", "fg_stdev")
  
  # Plot the data the same way as the teams were done
  ggplot(krig_fg_df, aes(x = x, y = y, fill = fg_pred)) +
    geom_raster(interpolate = TRUE) +
    draw_court() +
    scale_fill_viridis_c(labels = scales::percent,limits = c(0.20, 0.60),oob = scales::squish) +
    coord_fixed(xlim = c(0, 45), ylim = c(-25, 25)) +
    labs(title = paste(player, "- FG% Kriging (min FGA = 3)"),
         x = "Distance from Basket (ft)",
         y = "Court Width (ft)",
         fill = "FG%") +
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(), plot.margin = margin(t = 1, r = 0, b = 0, l = 3))
}
```

Kriging has 2 major limitations: sparse data and giant data. The giant data problem is avoided for this case by filtering out all of the free throws and plays that did not result in a shot attempt from the large play-by-play dataset, but slight manual adjustments had to be made for sparse data:

\begin{columns}
\begin{column}{0.5\textwidth}

This plot of Stephen Curry's individual FG\% Kriging from last season makes it clear that having too little data can be a problem as well. Not only are the bins that have enough shots to analyze somewhat blocky, but the areas where there are no data have been pulled towards the global mean which is obviously unreasonable given the context.

This emphasizes that Kriging is far better at interpolation than extrapolation.

\end{column}
\begin{column}{0.5\textwidth}
```{r curry-fg-plot, fig.width=2.8, fig.height=2.4, warning=FALSE, results='hide'}
plot_fg_pct_player("Stephen Curry")
```
\end{column}
\end{columns}

## Conclusions

\begin{columns}
\begin{column}{0.5\textwidth}
\nocite{*}
A plot of the residuals from the predictions against the distance from the basket of the shots gives some insights about how Kriging outputs tend to work - the model underpredicted the almost outlier-like high percentages on dunks and layups near the basket, was most wild for the midrange where the data is far more sparse and inconsistent, and then was fairly tight and accurate for three point shots.

This means that NBA teams could have the most confidence in conclusions related to threes but should perhaps be careful when making broad conclusions about midrange shots based on the Kriging technique.

\end{column}
\begin{column}{0.5\textwidth}
```{r nba-acc-plot, fig.width=3, fig.height=2.5, warning=FALSE, results='hide'}
# Call the helper function for the grid and bins back to normal size
pred_grid <- make_pred_grid()
bins <- team_shots %>%
  mutate(x_bin = floor(x / bin_size) * bin_size + bin_size / 2,
    y_bin = floor(y / bin_size) * bin_size + bin_size / 2) %>%
  group_by(x_bin, y_bin) %>%
  
  # Calculate the makes, attempts, and field goal percentage of each bin
  summarise(
    attempts = n(),
    makes = sum(score_value > 0),
    fg_pct = makes / attempts,
    .groups = "drop") %>%
  
  # Make sure there is enough attempts in each bin (easy for whole NBA)
  filter(attempts >= min_attempts) %>%
  
  # Calculate the distance and point value of each bin
  mutate(
    dist_from_rim = sqrt((x_bin - rim_x)^2 + y_bin^2),
    is_three = dist_from_rim >= three_pt_r | (abs(y_bin) >= 22 & x_bin <= corner_x),
    in_paint = x_bin <= paint_x & abs(y_bin) <= paint_y,
    zone = case_when(
        is_three ~ "Three",
      in_paint ~ "Paint",
      TRUE     ~ "Midrange")) %>%
  as.data.frame()
fg_bins <- bins %>% select(x_bin, y_bin, fg_pct) %>% as.data.frame()
coordinates(fg_bins) <- ~x_bin+y_bin

# Run Kriging on the league wide field goal percentage data
krig_fg <- suppressMessages(suppressWarnings(autoKrige(fg_pct ~ 1, input_data = fg_bins, new_data = pred_grid)))

# Make a dataframe of the Kriging output
krig_fg_df <- as.data.frame(krig_fg$krige_output)
names(krig_fg_df) <- c("x", "y", "fg_pred", "fg_var", "fg_stdev")

# Snap the bins to align them again
bins %>%
  mutate(
    x_snap = round((x_bin - 0.5)) + 0.5,
    y_snap = round((y_bin - 0.5)) + 0.5
  ) %>%
  left_join(krig_fg_df, by = c("x_snap" = "x", "y_snap" = "y")) %>%
  filter(!is.na(fg_pred)) %>%
  
  # Calculate the residual at each bin
  mutate(residual = fg_pct - fg_pred) %>%
  
  # Plot the reisudals against shot distance
  ggplot(aes(x = dist_from_rim, y = residual, color = zone, size = attempts)) +
  geom_point(alpha = 0.75) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey40", linewidth = 0.6) +
  
  # Color the data by shot type
  scale_color_manual(values = c("Paint" = "#C0392B", "Midrange" = "#F39C12", "Three" = "#2471A3")) +
  scale_size_continuous(range = c(1, 5), guide = "none") +
  scale_y_continuous(labels = scales::percent) +
  
  # Add labels
  labs(title = "NBA - Residuals vs. Distance",
       x = "Distance from Basket (ft)",
       y = "Actual - Predicted FG%",
       color = "Zone") +
  
  # Adjust the positioning on the plot
  theme_minimal(base_size = 7) +
  theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = 2))
```
\end{column}
\end{columns}

