---
title: "Kriging in Sports"
subtitle: "CMDA-4654 Project 2"
author: "Group 1 | Baker Dean, Toshaani Dewan, Nick Fitti, Nico Ottobre, Margaret Smith"
date: "April 2026"
fontsize: 9pt
classoption: "aspectratio=169"
output: 
  beamer_presentation: 
    fonttheme: professionalfonts
    highlight: kate
    theme: Rochester
urlcolor: blue
header-includes:
- \definecolor{VTmaroon}{HTML}{861F41}
- \usecolortheme[named=VTmaroon]{structure}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{bbm}
- \usepackage{cancel}
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{bm}
- \usepackage{enumitem}
- \usepackage{underscore}
- \setlist[itemize]{label=\textbullet}
---
```{r setup, include=FALSE}
#Libraries
library(knitr)
library(dplyr)
library(ggplot2)
library(hoopR)
library(progressr)
library(gstat)
library(sp)
library(automap)
library(patchwork)
library(scales)
library(sportyR)

# Setup
knitr::opts_chunk$set(echo = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")

def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  paste0("\n \\", "footnotesize","\n\n", x, "\n\n \\normalsize")
})
```

```{r load-nba-data, cache=TRUE, include=FALSE}
# Read in NBA play-by-play data for the 2025-26 season
pbp <- load_nba_pbp(seasons = 2026)
player_box <- load_nba_player_box(seasons = 2026)

# Read in a dictionary of player names and their IDs for the NBA data
player_names <- player_box %>% distinct(athlete_id, athlete_display_name)
```

```{r, include=FALSE}
# Unprinted Code for NBA Example

# Constants
rim_x <- 5.25
three_pt_r <- 23.75
corner_x <- rim_x + sqrt(three_pt_r^2 - 22^2)
paint_x <- 19
paint_y <- 8
bin_size <- 2
min_attempts <- 10

# Function to draft the lines of a basketball halfcourt
draw_court <- function() {
  
  # Define the angle of the three point line
  angle_top <- atan2(22, corner_x - rim_x)
  angle_bot <- atan2(-22, corner_x - rim_x)
  
  # Build the three point line using arc points
  arc_points <- data.frame(
    x = rim_x + three_pt_r * cos(seq(angle_top, angle_bot, length.out = 1000)),
    y = three_pt_r * sin(seq(angle_top, angle_bot, length.out = 1000))
  )
  
  # Make a list of the output elements
  list(
    # Three point arc
    geom_path(data = arc_points, aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # Corner three lines
    geom_path(data = data.frame(x = c(0, corner_x), y = c(22, 22)),
              aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    geom_path(data = data.frame(x = c(0, corner_x), y = c(-22, -22)),
              aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # Paint
    geom_path(data = data.frame(x = c(0, 19, 19, 0, 0), y = c(8, 8, -8, -8, 8)),
              aes(x = x, y = y), inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # FT circle solid
    geom_path(data = data.frame( x = 19 + 6 * cos(seq(-pi/2, pi/2, length.out = 500)),
                                 y = 6 * sin(seq(-pi/2, pi/2, length.out = 500))), aes(x = x, y = y),
              inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # FT circle dashed
    geom_path(data = data.frame(x = 19 + 6 * cos(seq(pi/2, 3*pi/2, length.out = 500)),
                                y = 6 * sin(seq(pi/2, 3*pi/2, length.out = 500))), aes(x = x, y = y),
              inherit.aes = FALSE, color = "black", linewidth = 0.6, linetype = "dashed"),
    # Restricted arc
    geom_path(data = data.frame(x = rim_x + 4 * cos(seq(-pi/2, pi/2, length.out = 500)),
                                y = 4 * sin(seq(-pi/2, pi/2, length.out = 500))), aes(x = x, y = y),
              inherit.aes = FALSE, color = "black", linewidth = 0.6),
    # Rim
    geom_path(data = data.frame(x = rim_x + 0.75 * cos(seq(0, 2*pi, length.out = 500)),
                                y = 0.75 * sin(seq(0, 2*pi, length.out = 500))), aes(x = x, y = y),
              inherit.aes = FALSE, color = "darkorange2", linewidth = 0.8))
}

# Data Cleaning
shots <- pbp %>%
  
  # Filter to plays that have shooting coordinates
  filter(shooting_play == TRUE,
         !is.na(coordinate_x),
         !is.na(coordinate_y),
         # Remove free throws
         !grepl("Free Throw", type_text),
         #Filter to the regular season
         season_type == 2) %>%
    
  # Translate from player IDs to names
  left_join(player_names, by = c("athlete_id_1" = "athlete_id")) %>%
    
  #Flip the data into being only one half of the court
  mutate(team_name = ifelse(team_id == home_team_id,
         paste(home_team_name, home_team_mascot), paste(away_team_name, away_team_mascot)),
    x = ifelse(coordinate_x > 0, -(coordinate_x - 46.75), coordinate_x + 46.75),
    y = ifelse(coordinate_x > 0, -coordinate_y, coordinate_y)) %>%
  
  # Get rid of heaves from beyond half court
  filter(x <= 45) %>%
    
  # Narrow the columns again
  select(athlete_display_name, team_name, score_value, x, y)

  
# Helper function to make the prediction grid
make_pred_grid <- function() {
  
  # Make a grid of squares across the court
  pred_grid <- expand.grid(
    x = seq(0.5, 45.5, by = 1),
    y = seq(-24.5, 25.5, by = 1)) %>%
    as.data.frame() %>%
    
    # Calculate the distance from the rim of each point
    mutate(dist_from_rim = sqrt((x - rim_x)^2 + y^2)) %>%
    
    # Filter to points that are within 5 feet of the three point line
    filter(dist_from_rim <= three_pt_r + 5) %>%
    select(x, y) %>%
    as.data.frame()
  
  # Make the grid itself
  coordinates(pred_grid) <- ~x+y
  gridded(pred_grid) <- TRUE
  pred_grid
}

# Helper function to make the bins for a team
make_bins <- function(team, bin_size = 2, min_attempts = 10) {
  
  # Filter to only shot attempts from a specified team
  team_shots %>%
    filter(team_name == team) %>%
    
    # Create 2 ft x 2 ft bins across the court
    mutate(
      x_bin = floor(x / bin_size) * bin_size + bin_size / 2,
      y_bin = floor(y / bin_size) * bin_size + bin_size / 2) %>%
    group_by(x_bin, y_bin) %>%
    
    # Define the attempts, makes, and field goal percent of each bin
    summarise(
      attempts = n(),
      makes = sum(score_value > 0),
      fg_pct = makes / attempts,
      .groups = "drop") %>%
    
    # Ensure the bin has enough attempts to be meaningful
    filter(attempts >= min_attempts) %>%
    
    # Add characteristics of each bin like point value
    mutate(
      dist_from_rim = sqrt((x_bin - rim_x)^2 + y_bin^2),
      is_three = dist_from_rim >= three_pt_r | (abs(y_bin) >= 22 & x_bin <= corner_x),
      in_paint = x_bin <= paint_x & abs(y_bin) <= paint_y,
      zone = case_when(
        is_three ~ "Three",
        in_paint ~ "Paint",
        TRUE     ~ "Midrange")) %>%
    as.data.frame()
}
```

## Data Introduction

Points per shot is one of the biggest current topics of discussion in basketball analytics. It is calculated using two factors: a team or players chance of making a shot from a position on the floor and the point of value of a make from that position. Performing analysis on this topic requires gathering both of those values which can be done using the `hoopr` package.

```{r, echo=TRUE, eval=FALSE}
library(hoopR)

pbp <- load_nba_pbp(seasons = 2026)
player_box <- load_nba_player_box(seasons = 2026)

player_names <- player_box %>%
  distinct(athlete_id, athlete_display_name)

shots <- pbp %>%
  filter(shooting_play == TRUE,
         !is.na(coordinate_x), !is.na(coordinate_y), !grepl("Free Throw", type_text),
         season_type == 2) %>%
  select(athlete_id_1, coordinate_x, coordinate_y, score_value) %>%
  left_join(player_names, by = c("athlete_id_1" = "athlete_id")) %>%
  mutate(
    x = ifelse(coordinate_x > 0,
               -(coordinate_x - 46.75),
               coordinate_x + 46.75),
    y = ifelse(coordinate_x > 0, -coordinate_y, coordinate_y)) %>%
  filter(x <= 45) %>%
  select(athlete_display_name, score_value, x, y)
```

## Data Example

The data has been loaded and saved as csv for ease of access: \href{https://docs.google.com/spreadsheets/d/1uBDuThoKEiOK03TE7cipW_oFBo6iOxBon5RDeMG9lUI/edit?usp=sharing}{Click here} to access shots.csv. 
\vspace{0.15cm}
\begin{columns}
\begin{column}{0.5\textwidth}

Preview of the data
\vspace{0.1cm}
```{r}
# Look at the first 20 rows of the shots data
shots %>% 
  head(20) %>%
  knitr::kable(format = "latex", booktabs = TRUE)
```

\end{column}
\begin{column}{0.5\textwidth}

\quad{}\quad{}\quad{}NBA 2025-26 Regular Season Heatmap
\vspace{0.1cm}
```{r, fig.width=4, fig.height=2.33}
# Make a plot of the NBA heatmap from the 2025-26 season
ggplot(shots, aes(x = x, y = y)) +
  stat_density_2d(aes(fill = after_stat(density)),geom = "raster",contour = FALSE,alpha = 0.9) +
  
  # Draw the court lines
  draw_court() +
  scale_fill_viridis_c() +
  
  #Establish the boundaries
  coord_fixed(xlim = c(0, 45),ylim = c(-25, 25),expand = FALSE,clip = "off") +
  
  # Add axis labels
  labs(title = "NBA Shot Locations",x = "Distance from Basket (ft)",y = "Court Width (ft)",fill = "Density") +
  
  # Adjust the plot positioning
  theme_minimal(base_size = 7) + 
  theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = -50))
```
\end{column}
\end{columns}

## Kriging Motivation

It is clear from the shot density layout on the previous slide that there are two main areas that teams take shots from:

\begin{itemize}
  \item the three point line
  \item the rim
\end{itemize}

This makes sense given that shots at the rim are likely to be high percentage and shots from three point range are worth more points. The middle area between these two hot zones is fairly empty, and the league has been trending to make midrange jumpshots even less a part of offenses in the Modern Era. All of this evolution is in the pursuit of offensive efficiency which is where Kriging comes in.

Since the shot map is made up of x and y data, it can be treated similarly to a geostatistics field from more typical Kriging -  it has high density areas and low gaps with a large number of very individualized spots. Kriging can be used to smooth over these less commonly attempted areas and give a picture of not only where teams shoot but how well they perform from these areas.

Combining those two areas will mean that Kriging can make a points per shot estimation from every location on the floor which can then be used to check if teams are running their offenses to peak efficiency.

## Attempt Frequency Kriging

A typical picture of attempt frequency for a basketball team would consist of colored boxes or hexes that represent volume from each area of the court, but a custom Kriging can smooth this out into a continuous map using two foot by two foot bins.

```{r plot-attempts}
# Get the shots for a specific team
team_shots <- shots %>% select(team_name, score_value, x, y)

# Define a function to perform Kriging on the shot attempts of a team
plot_attempts <- function(team) {
  
  # Use helper functions to build correct bins
  pred_grid <- make_pred_grid()
  bins <- make_bins(team)
  attempt_bins <- bins %>% select(x_bin, y_bin, attempts) %>% as.data.frame()
  coordinates(attempt_bins) <- ~x_bin+y_bin
  
  # Run autoKrige on the shot attempts data
  krig_attempts <- suppressWarnings(autoKrige(attempts ~ 1, input_data = attempt_bins, new_data = pred_grid))
  
  # Form a dataframe from the Kriging output
  krig_attempts_df <- as.data.frame(krig_attempts$krige_output)
  names(krig_attempts_df) <- c("x", "y", "attempts_pred", "attempts_var", "attempts_stdev")
  krig_attempts_df <- krig_attempts_df %>% mutate(attempts_pred = pmax(0, attempts_pred))
  
  # Plot the output using helper functions
  ggplot(krig_attempts_df, aes(x = x, y = y, fill = attempts_pred)) +
    geom_raster(interpolate = TRUE) +
    draw_court() +
    scale_fill_viridis_c() +
    coord_fixed(xlim = c(0, 35), ylim = c(-25, 25)) +
    labs(title = paste(team, "- Shot Attempts"),
         x = "Distance from Basket (ft)",
         y = "Court Width (ft)",
         fill = "Attempts") +
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = -3))
}
```
\begin{columns}
\begin{column}{0.5\textwidth}

```{r bos-attempts-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_attempts("Boston Celtics")
```

\end{column}
\begin{column}{0.5\textwidth}

```{r det-attempts-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_attempts("Detroit Pistons")
```

\end{column}
\end{columns}

## FG\% Kriging

While the Kriging for shot attempts is somewhat trivial, it is extremely useful for field goal percentage. This statistic leads to a map that has continuous values across all the bins and the maps will differ more greatly team-to-team.

```{r plot-percentage}
# Define a function to perform Kriging on the field goal percentage of a team
plot_fg_pct <- function(team) {
  
  # Call helper functions to build the correct bins
  pred_grid <- make_pred_grid()
  bins <- make_bins(team)
  fg_bins <- bins %>% select(x_bin, y_bin, fg_pct) %>% as.data.frame()
  coordinates(fg_bins) <- ~x_bin+y_bin
  
  # Call autoKrige on the field goal percentage data
  krig_fg_auto <- suppressMessages(suppressWarnings(autoKrige(fg_pct ~ 1, input_data = fg_bins, new_data = pred_grid)))
  
  # Create a df from the Kriging output
  krig_fg_df <- as.data.frame(krig_fg_auto$krige_output)
  names(krig_fg_df) <- c("x", "y", "fg_pred", "fg_var", "fg_stdev")
  
  # Return both a plot of the output and the data itself
  return(list(
    plot = ggplot(krig_fg_df, aes(x = x, y = y, fill = fg_pred)) +
      geom_raster(interpolate = TRUE) +
      draw_court() +
      scale_fill_viridis_c(labels = scales::percent,
                           limits = c(0.20, 0.60),
                           oob = scales::squish) +
      coord_fixed(xlim = c(0, 35), ylim = c(-25, 25)) +
      labs(title = paste(team, "- FG%"),
           x = "Distance from Basket (ft)",
           y = "Court Width (ft)",
           fill = "FG%") +
      theme_minimal(base_size = 7) +
      theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = -3)),
    data = krig_fg_df,
    krig = krig_fg_auto))
}
```
\begin{columns}
\begin{column}{0.5\textwidth}

```{r bos-percentage-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
bos_fg <- plot_fg_pct("Boston Celtics")
bos_fg$plot
```

\end{column}
\begin{column}{0.5\textwidth}

```{r det-percentage-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
det_fg <- plot_fg_pct("Detroit Pistons")
det_fg$plot
```

\end{column}
\end{columns}

## Points Per Shot

As mentioned previously, points per shot is a metric that is used to determine the quality of a type of shot attempt. Now that the field goal percentage Kriging has supplied a likelihood of making a shot at any location on the floor, it is easy to take it a step further into point values of the shots.

```{r}
# Define a function to perform Kriging on the points per shot data
plot_pps <- function(fg_data, team = "") {
  
  # Define the number of points each shot is worth
  fg_data %>%
    mutate(dist_from_rim = sqrt((x - rim_x)^2 + y^2),
      is_three = dist_from_rim >= three_pt_r | (abs(y) >= 22 & x <= corner_x),
      point_value = ifelse(is_three, 3, 2),pps = fg_pred * point_value) %>%
    
    # Plot the data using helper functions
    ggplot(aes(x = x, y = y, fill = pps)) +
    geom_raster(interpolate = TRUE) +
    draw_court() +
    scale_fill_viridis_c(limits = c(0.5, 2.0), oob = scales::squish) +
    coord_fixed(xlim = c(0, 35), ylim = c(-25, 25)) +
    labs(title = paste(team, "- Points per Shot"),
         x = "Distance from Basket (ft)",
         y = "Court Width (ft)",
         fill = "PPS") +
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = -3))
}
```
\begin{columns}
\begin{column}{0.5\textwidth}

```{r bos-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(bos_fg$data, "Boston Celtics")
```

\end{column}
\begin{column}{0.5\textwidth}

```{r det-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(det_fg$data, "Detroit Pistons")
```

\end{column}
\end{columns}

## Non-Kriging Approach

So the question becomes "How do these smooth maps help NBA teams improve their offenses?"

One of the main advantages that these plots provide is the way they make it much easier to full picture of slots across the floor. As a comparison, here would be a more standard look at a teams percentages and shot volumes from Synergy Sports. 

\begin{columns}
\begin{column}{0.5\textwidth}
The plot makes it hard to tell whether there's smaller deviations within the large group and doesn't calculate the point values of the shots (i.e. shooting 50\% from 2 is the same as 33\% from 3). Thus it is generally not very insightful about which shots are the most efficient, only offering general statistics about which areas the team has shot the most accurately from.
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=\linewidth]{"Synergy Shot Chart.png"}
\end{column}
\end{columns}

## Kriging Comparison

A smooth chart from Kriging much better captures the specific peaks and valleys of the shooting percentage and factoring in the point values gives real, actionable insights. Combining this with the shots attempted map can make it easy for a coach without a statistical background to see mismatches in shot value and frequency without complex interpretation.

Running through every NBA team can make some interesting comparisons:

```{r, eval=FALSE, echo=TRUE}
plot_team_impact <- function(team_name) {
   fg_obj <- plot_fg_pct(team_name)
   attempts_plot <- plot_attempts(team_name)
   pps_plot <- plot_pps(fg_obj$data)
   attempts_plot | pps_plot
}

nba_teams <- sort(unique(team_shots$team_name))

for (tm in nba_teams) {
   print(plot_team_impact(tm))
}
```

## Team Takeaways

```{r, fig.width=3, fig.height=3, warning=FALSE, results='hide'}
# Interesting comparisons
cle <- plot_fg_pct("Cleveland Cavaliers")$data


# Interesting side-to-side plots
ind <- plot_fg_pct("Indiana Pacers")$data
den <- plot_fg_pct("Denver Nuggets")$data
```

Some teams could gain insights by looking at their charts side-by-side, like the Cleveland Cavaliers

\begin{columns}
\begin{column}{0.5\textwidth}
```{r cle-attempts-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_attempts("Cleveland Cavaliers")
```
\end{column}
\begin{column}{0.5\textwidth}
```{r cle-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(cle, "Cleveland Cavaliers")
```
\end{column}
\end{columns}

## Points per Shot Patterns

Some teams also have clear patterns that favor shooting from one side of the floor

\begin{columns}
\begin{column}{0.5\textwidth}
```{r ind-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(ind, "Indiana Pacers")
```
\end{column}
\begin{column}{0.5\textwidth}
```{r den-pps-plot, fig.width=2.5, fig.height=2.5, warning=FALSE, results='hide'}
plot_pps(den, "Denver Nuggets")
```
\end{column}
\end{columns}

## Accuracy of Predictions

Of course, the predicted maps are only useful if they are accurate. So how well did Kriging do with smoothing the shooting data?

```{r}
# Define a function to show the accuracy of the Kriging predictions
plot_accuracy <- function(team) {
  
  # Call the helper functions
  bins <- make_bins(team)
  krig_fg_df <- plot_fg_pct(team)$data
  
  # Adjust the bins to match the real data
  bins %>%
    mutate(x_snap = round((x_bin - 0.5)) + 0.5, y_snap = round((y_bin - 0.5)) + 0.5) %>%
    left_join(krig_fg_df, by = c("x_snap" = "x", "y_snap" = "y")) %>%
    filter(!is.na(fg_pred)) %>%
    
    # Make a plot of the accuracy
    ggplot(aes(x = fg_pct, y = fg_pred, color = zone, size = attempts)) +
    geom_point(alpha = 0.75) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.6) +
    
    # Color code the plot by shot type
    scale_color_manual(values = c("Paint" = "#C0392B", "Midrange" = "#F39C12", "Three" = "#2471A3")) +
    scale_size_continuous(range = c(1, 5), guide = "none") +
    scale_x_continuous(labels = scales::percent, limits = c(0.15, 0.85)) +
    scale_y_continuous(labels = scales::percent, limits = c(0.15, 0.85)) +
    
    # Add labels
    labs(title = paste(team, "- Kriging Accuracy"), x = "Actual FG%", y = "Kriging Predicted FG%", color = "Zone") +
    
    # Adjust the positioning of the plot
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = 2))
}
```

\begin{columns}
\begin{column}{0.5\textwidth}
```{r bos-acc-plot, fig.width=2.75, fig.height=2.4, warning=FALSE, results='hide'}
plot_accuracy("Boston Celtics")
```
\end{column}
\begin{column}{0.5\textwidth}
```{r det-acc-plot, fig.width=2.75, fig.height=2.4, warning=FALSE, results='hide'}
plot_accuracy("Detroit Pistons")
```
\end{column}
\end{columns}

## Kriging Limitations

```{r}
# Define a function to plot the field goal percentage Kriging of a specific player
plot_fg_pct_player <- function(player) {
  
  # Define a fully expanded half court grid for the players
  pred_grid <- expand.grid(x = seq(0.5, 45.5, by = 1),y = seq(-24.5, 25.5, by = 1)) %>%
    as.data.frame() %>% select(x, y) %>% as.data.frame()
  coordinates(pred_grid) <- ~x+y
  gridded(pred_grid) <- TRUE
  
  # Re-define the bins to match this new grid
  fg_bins <- shots %>%
    filter(athlete_display_name == player) %>%
    mutate(x_bin = floor(x / bin_size) * bin_size + bin_size / 2,
      y_bin = floor(y / bin_size) * bin_size + bin_size / 2) %>%
    group_by(x_bin, y_bin) %>%
    summarise(
      attempts = n(),
      makes = sum(score_value > 0),
      fg_pct = makes / attempts,
      .groups = "drop") %>%
    filter(attempts >= 3) %>%
    as.data.frame()
  coordinates(fg_bins) <- ~x_bin+y_bin
  
  # Run Kriging on the individual field goal percentage data
  krig_fg <- suppressMessages(suppressWarnings(autoKrige(fg_pct ~ 1, input_data = fg_bins, new_data = pred_grid)))
  
  # Form a data frame on the Kriging output
  krig_fg_df <- as.data.frame(krig_fg$krige_output)
  names(krig_fg_df) <- c("x", "y", "fg_pred", "fg_var", "fg_stdev")
  
  # Plot the data the same way as the teams were done
  ggplot(krig_fg_df, aes(x = x, y = y, fill = fg_pred)) +
    geom_raster(interpolate = TRUE) +
    draw_court() +
    scale_fill_viridis_c(labels = scales::percent,limits = c(0.20, 0.60),oob = scales::squish) +
    coord_fixed(xlim = c(0, 45), ylim = c(-25, 25)) +
    labs(title = paste(player, "- FG% Kriging (min FGA = 3)"),
         x = "Distance from Basket (ft)",
         y = "Court Width (ft)",
         fill = "FG%") +
    theme_minimal(base_size = 7) +
    theme(panel.grid = element_blank(), plot.margin = margin(t = 1, r = 0, b = 0, l = 3))
}
```

Kriging has 2 major limitations: sparse data and giant data. The giant data problem is avoided for this case by filtering out all of the free throws and plays that did not result in a shot attempt from the large play-by-play dataset, but slight manual adjustments had to be made for sparse data:

\begin{columns}
\begin{column}{0.5\textwidth}

This plot of Stephen Curry's individual FG\% Kriging from last season makes it clear that having too little data can be a problem as well. Not only are the bins that have enough shots to analyze somewhat blocky, but the areas where there are no data have been pulled towards the global mean which is obviously unreasonable given the context.

This emphasizes that Kriging is far better at interpolation than extrapolation.

\end{column}
\begin{column}{0.5\textwidth}
```{r curry-fg-plot, fig.width=2.8, fig.height=2.4, warning=FALSE, results='hide'}
plot_fg_pct_player("Stephen Curry")
```
\end{column}
\end{columns}

## Conclusions

\begin{columns}
\begin{column}{0.5\textwidth}

A plot of the residuals from the predictions against the distance from the basket of the shots gives some insights about how Kriging outputs tend to work - the model underpredicted the almost outlier-like high percentages on dunks and layups near the basket, was most wild for the midrange where the data is far more sparse and inconsistent, and then was fairly tight and accurate for three point shots.

This means that NBA teams could have the most confidence in conclusions related to threes but should perhaps be careful when making broad conclusions about midrange shots based on the Kriging technique.

\end{column}
\begin{column}{0.5\textwidth}
```{r nba-acc-plot, fig.width=3, fig.height=2.5, warning=FALSE, results='hide'}
# Call the helper function for the grid and bins back to normal size
pred_grid <- make_pred_grid()
bins <- team_shots %>%
  mutate(x_bin = floor(x / bin_size) * bin_size + bin_size / 2,
    y_bin = floor(y / bin_size) * bin_size + bin_size / 2) %>%
  group_by(x_bin, y_bin) %>%
  
  # Calculate the makes, attempts, and field goal percentage of each bin
  summarise(
    attempts = n(),
    makes = sum(score_value > 0),
    fg_pct = makes / attempts,
    .groups = "drop") %>%
  
  # Make sure there is enough attempts in each bin (easy for whole NBA)
  filter(attempts >= min_attempts) %>%
  
  # Calculate the distance and point value of each bin
  mutate(
    dist_from_rim = sqrt((x_bin - rim_x)^2 + y_bin^2),
    is_three = dist_from_rim >= three_pt_r | (abs(y_bin) >= 22 & x_bin <= corner_x),
    in_paint = x_bin <= paint_x & abs(y_bin) <= paint_y,
    zone = case_when(
        is_three ~ "Three",
      in_paint ~ "Paint",
      TRUE     ~ "Midrange")) %>%
  as.data.frame()
fg_bins <- bins %>% select(x_bin, y_bin, fg_pct) %>% as.data.frame()
coordinates(fg_bins) <- ~x_bin+y_bin

# Run Kriging on the league wide field goal percentage data
krig_fg <- suppressMessages(suppressWarnings(autoKrige(fg_pct ~ 1, input_data = fg_bins, new_data = pred_grid)))

# Make a dataframe of the Kriging output
krig_fg_df <- as.data.frame(krig_fg$krige_output)
names(krig_fg_df) <- c("x", "y", "fg_pred", "fg_var", "fg_stdev")

# Snap the bins to align them again
bins %>%
  mutate(
    x_snap = round((x_bin - 0.5)) + 0.5,
    y_snap = round((y_bin - 0.5)) + 0.5
  ) %>%
  left_join(krig_fg_df, by = c("x_snap" = "x", "y_snap" = "y")) %>%
  filter(!is.na(fg_pred)) %>%
  
  # Calculate the residual at each bin
  mutate(residual = fg_pct - fg_pred) %>%
  
  # Plot the reisudals against shot distance
  ggplot(aes(x = dist_from_rim, y = residual, color = zone, size = attempts)) +
  geom_point(alpha = 0.75) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey40", linewidth = 0.6) +
  
  # Color the data by shot type
  scale_color_manual(values = c("Paint" = "#C0392B", "Midrange" = "#F39C12", "Three" = "#2471A3")) +
  scale_size_continuous(range = c(1, 5), guide = "none") +
  scale_y_continuous(labels = scales::percent) +
  
  # Add labels
  labs(title = "NBA - Residuals vs. Distance",
       x = "Distance from Basket (ft)",
       y = "Actual - Predicted FG%",
       color = "Zone") +
  
  # Adjust the positioning on the plot
  theme_minimal(base_size = 7) +
  theme(panel.grid = element_blank(), plot.margin = margin(t = 2, r = 0, b = 0, l = 2))
```
\end{column}
\end{columns}

## References

\href{https://www.paulamoraga.com/book-spatial/kriging.html}{https://www.paulamoraga.com/book-spatial/kriging.html}

\href{https://www.publichealth.columbia.edu/research/population-health-methods/kriging-interpolation}{https://www.publichealth.columbia.edu/research/population-health-methods/kriging-interpolation}

\href{https://www.youtube.com/watch?v=dfSMGJFRL1U}{https://www.youtube.com/watch?v=dfSMGJFRL1U}

\href{https://www.youtube.com/watch?v=ZRyiBKvqjJM}{https://www.youtube.com/watch?v=ZRyiBKvqjJM}

\href{https://link.springer.com/article/10.1007/s00180-024-01564-4}{https://link.springer.com/article/10.1007/s00180-024-01564-4}